---
title: "Aggregating over subperiods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aggregating over subperiods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

With the exception of `vignette("spatial-price-index")`, all the examples so far
have revolved around aggregating over the levels of a price index for each
time period. Although this is the core workflow in this package, it is also
useful to be able to aggregate over the time periods for each level of an index
to turn a monthly or quarterly index into an annual index.

Let's modify the example in `vignette("adjust-weights")` by adding an additional
eight quarters to make three years of data.

```{r}
set.seed(54321)

library(piar)

# Make an aggregation structure.
#                  1
#      |-----------|-----------|
#      11          12          13
#  |---+---|   |---+---|   |---+---|
# 111     121 121     122 131     132

pias <- data.frame(
  level1 = rep(1, 12),
  level2 = rep(c(11, 12, 13), each = 4),
  level3 = rep(c(111, 112, 121, 122, 131, 132), each = 2),
  ea = sprintf("B%02d", 1:12),
  weight = 1:12
) |>
  as_aggregation_structure()

pias

# Make elemental indexes over 3 years and aggregate.
quarterly_index <- matrix(
  runif(12 * 12, 0.4, 1.2),
  nrow = 12,
  dimnames = list(sprintf("B%02d", 1:12), paste0("Q", 1:12))
) |>
  as_index() |>
  aggregate(pias)

head(quarterly_index)
```

The conventional way to turn a quarterly arithmetic index into an annual one is
to take the (unweighted) arithmetic mean of the index values over each year and
rebase to a new base year.

```{r}
annual_index <- chain(quarterly_index) |>
  mean(window = 4)

annual_index |>
  rebase("Q1") |>
  head()
```

It's worth not that, at least with an arithmetic index, the aggregation
properties of the index continue to hold after price-updating the weights.

```{r}
annual_pias <- pias |>
  update(annual_index, period = "Q1")

annual_index <- annual_index |>
  rebase("Q1")

annual_index |>
  aggregate(annual_pias) |>
  all.equal(annual_index)
```

This means that, for example, the workflow in `vignette("contributions")` for
determining, say, the quarter-over-quarter contribution of the level 2 indexes
to top-level index remains the same.

```{r}
annual_index |>
  unchain() |>
  set_contrib_from_index() |>
  aggregate(cut(annual_pias, 2)) |>
  contrib()
```